---
format: html
bibliography: references.bib
echo: false
---

## Bottlenose dolphins (*Tursiops truncatus*) kill harbor porpoises (*Phocoena phocoena*): evidence from the central California stranding record, 2007-2023.

Authors: William Keener, Marc Webber, Padraig Duignan, Jackson Vanfleet-Brown


```{r}
#| output: false

library(readxl)
library(readr)
library(ggOceanMaps)
library(ggspatial)
library(RColorBrewer)
library(tidyverse)

#raw data files
rawFiles <- list.files("data/raw", full.names = TRUE)
#omit the unneeded spreadsheet in data/raw
rawFiles <- rawFiles[-1]
#sheets needed in each data file
rawSheets <- c(1,1)
#lookup table with probability of porpocide for every case
lookup <- read_csv(file="data/derived/Porpicide_Lookup.csv", col_types = "cf")
#vars wanted (note these vary nominally in each spreadsheet)
varsWanted <- c("TMMC_ID","Age_Class","Sex","Strand_Date","Strand_County", "County", "Latitude", "Longitude")
  
getData <- function(raw, vars) {
  result <- raw %>% 
    select(any_of(vars)) %>% 
    left_join(lookup, by = "TMMC_ID") %>%
    mutate(year = year(Strand_Date))%>%
    mutate(yearf = as.factor(year)) %>%
    mutate(month = as.factor(month(Strand_Date))) %>%
    mutate(Latitude = as.numeric(Latitude)) %>%
    mutate(Longitude = as.numeric(Longitude)) %>%
    mutate(`CONFIDENCE  OF PORPICIDE` =  fct_relevel(`CONFIDENCE  OF PORPICIDE`, c("CONFIRMED", "PROBABLE", "SUSPECT", "UNLIKELY"))) %>%
    mutate(
      agef = case_when(
        Age_Class == "Adult" ~ "Adult",
        Age_Class == "Subadult" ~ "Subadult",
        Age_Class == "Juvenile" ~ "Juvenile",
        Age_Class == "Calf" ~ "Calf",
        .default = "Other"
        )
      ) %>%
    mutate(agef = as.factor(agef))
  names(result) <- c("id", "age", "sex", "date", "county", "lat", "lon", "prob", "year", "yearf", "month", "agef")
  levels(result$month) <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  #replacing non-trauma cases (NA because not reviewed by B.C.) with UNLIKELY
  result$prob <- replace_na(result$prob, "UNLIKELY")
  result$sex <- as.factor(replace_na(result$sex, "Unknown"))
  levels(result$agef) <- c("Adult", "Subadult", "Juvenile", "Calf", "Other")
  return(result)
}

myData <- map2(rawFiles, rawSheets, function(x,y) read_excel(path=x, sheet=y))
myData <- map(myData, function(x) getData(x, vars=varsWanted))
myData <- list_rbind(myData)

#define palette for plots
myPal <- brewer.pal(n=length(unique(myData$prob)), "YlOrRd")
names(myPal) <- rev(levels(myData$prob))
```

## Introduction

Harbor porpoise (*Phocoena phocoeana*) populations in California have grown at rates of approximately 6-10% since 2004, the post-gillnet bycatch period [@forney2021]. Despite this growth, there exist concern about emerging threats. One such concern is attack by bottlenose dolphins (*Tursiops truncatus*). 

A recognized phenomenon worldwide [e.g. @gross_indication_2020], instances of interspecific aggression were first noted in California at the time of the "unusual mortality event" (UME) of 2007-2009. Wilkin et al. [-@wilkin2012] examined stranded carcasses of harbor porpoises collected during this period and identified blunt force trauma as the most common known cause of death. Injuries in these cases suggested attack by bottlenose dolphins. Cotter et al. [-@cotter_porpicide_2012] observed attacks in Monterey Bay during this same period, providing further evidence of porpicide. Both authors indicated that these types of attacks peaked in the late summer and fall.

In this study we reviewed the harbor porpoise stranding record in central California from the time of the UME reported by Wilkin et al. [-@wilkin2012] to the present, estimating the probability of porpicide (i.e. attack by bottlenose dolphin) in each case.

## Methods

-   Stranding network
    -   range of search area and protocol for reporting and collecting carcasses
-   Pathology
    -   process for necropsy and assessing TME data
    -   process for estimating probability
        -   Confirmed Case: Tt rakes + Fractures + Hemorrhage +/- Capture Myopathy
        -   Probable Case: Fractures + Hemorrhage +/- Capture Myopathy
        -   Suspect Case: Hemorrhage +/- Capture Myopathy
        -   Unlikely Case: Sudden death without blunt or sharp force trauma.

## Results & Discussion

### Interannual variability

```{r}
#| output: false
myData %>%
  group_by(year) %>%
  count()

```

The number of cases each year depends on mortality and to a large extent on other factors, such as the probability that carcasses will float to shore and that they will be found. This causes variability year to year in the data set. There is a significant increase in cases coincident with the beginning of the UME in 2008 [@wilkin2012]. This is all shown in @fig-yearly-counts.

```{r}
#| label: fig-yearly-counts
#| fig-cap: "Yearly counts are variable due to a combination of influences."

myData %>%
  ggplot(aes(x=yearf))+
  geom_bar()+
  xlab("Year")+
  ylab("Count of cases")+
  theme(axis.text.x=element_text(angle=90))
  # scale_fill_manual(values=myPal)
```

Normalizing for interannual variability, we see that evidence of porpicide was most prevalent among strandings in 2015-2020. This trend appears to have reversed over the past 3 years as depicted in @fig-prop-porp.

```{r}
#| label: fig-prop-porp
#| fig-cap: "Incidence of porpicide appears to increase, then plateau 2015-2020, before decreasing again in recent years."
countData <- myData %>%
  count(year, prob)

allYr <- rep(2007:2023, each=4)
allProb <- rep(c("UNLIKELY", "SUSPECT", "PROBABLE", "CONFIRMED"), 17)
nullData <- data.frame(year=allYr, prob=allProb, n=0)
nullData <- anti_join(nullData, countData, by=c("year", "prob"))
countData %>%
  rbind(nullData) %>%
  ggplot(aes(x=as.factor(year), y=n, fill=prob))+
  geom_bar(stat="identity", position="fill")+
  labs(y="Proportion of cases",x="Year", fill="Estimated probability")+
  scale_fill_manual(values=myPal)+
  theme(axis.text.x = element_text(angle=90))
```

### Seasonal variability

As seen in @fig-monthly-counts there is seasonal cycle in strandings. Most occur in the summer. This is consistent with Wilkin et al. [-@wilkin2012] who observed the same patterns in stranding data from 1998-2010.

```{r}
#| label: fig-monthly-counts
#| fig-cap: "Cumulative number of strandings, by month, for the period 2007-2023. Shows most strandings occur during the Summer."

myData %>%
  ggplot(aes(x=month))+
  geom_bar() +
  labs(y="Total cases", x="Month", fill="Estimated probability")
```

Porpicide cases appear to peak later in the year, in the late-summer to fall. @fig-monthly-probs shows that this peak occurs July-October.  *Note: as a next step in this analysis, we could test for a statistically significant difference between the distribution of high-probability and low-probability porpicide cases.*

```{r}
#| label: fig-monthly-probs
#| fig-cap: "Proportion of strandings in each probability class, by month. These proportions are derived from the 18 year period Jan 2007- Dec 2023"
#| message: false

myData %>%
  group_by(year, month) %>%
  count(prob) %>% 
  ggplot() +
  geom_bar(aes(x=month, fill=prob), position="fill")+
  scale_fill_manual(values=myPal) +
  labs(x="Month",y="Proportion of cases", fill="Estimated probability")

```

### Spatial extent

```{r}
#| output: false

#make maps
pal = colorRampPalette(colors = c("red", "blue"))
spat <- drop_na(myData, lat, lon)
maps <- spat %>%
  nest(data=-prob) %>%
  mutate(plot = map2(data, prob, function(x, y) basemap(x) + geom_spatial_point(x, mapping = aes(x=lon, y = lat, color=yearf), crs = 4326) + scale_color_discrete(type = pal(19)) + labs(color="Year") + ggtitle(y))) %>%
  pull(plot, name = prob)

# basemap(spat) +
#   geom_spatial_point(spat, mapping = aes(x=lon, y = lat, color=yearf), crs = 4326)+
#   scale_color_discrete(type = pal(19)) +
#   facet_wrap(~prob)
```
The northward range expansion of bottlenose dolphins could drive a higher incidence of porpicide in northern regions through time. An initial look at the spatial data [@fig-spat-conf] does not reveal any obvious northward progression in confirmed porpicide cases.

```{r}
#| label: fig-spat-conf
#| fig-cap: "Spatial extent of confirmed porpicide cases does not show any chronological trends, such as a northward progression which may have been the effect of range expansion documented in California bottlenose dolphins."

maps$CONFIRMED
```

If confirmed cases are compared to unconfirmed cases, a weak northward trend may appear. The maximum latitude of all confirmed cases in a given year appears to increase through time, particularly if to all unlikely cases.

```{r}
#| label: fig-maxlat
#| fig-cap: "The maximum latitude of strandings exhibits a weak northwards trend for high-likelihood porpicide cases, especially when compared to low-likelihood cases."
#| message: false
#| warning: false
myData %>%
  # mutate(prob2=fct_other(prob, keep="UNLIKELY", other_level = "LIKELY")) %>%
  filter(prob=="CONFIRMED"|prob=="UNLIKELY") %>% 
  group_by(year, prob) %>%
  summarize(maxlat=max(lat)) %>%
  ggplot()+
  geom_point(aes(x=year, y=maxlat))+
  labs(y="Maximum Latitude", x="Year")+
  facet_wrap(~prob)
```

### Age/sex of porpicides

```{r}
#| label: fig-demography
#| fig-cap: "Adult females and juvenile males appear to be the most frequent targets among confirmed cases of porpicide."
myData %>%
  filter(sex != "Unknown") %>%
  ggplot(aes(x=agef, fill=prob))+
  geom_bar()+
  facet_grid(prob~sex)+
  xlab("Age Class") +
  scale_fill_manual(values=myPal)
```

## References

::: {#refs}
:::
